---
title: "Simulated Hospital Data"
author: "Andy Grogan-Kaylor (agrogan@umich.edu)"
date: "`r Sys.Date()`"
output: 
  slidy_presentation: 
    highlight: haddock
---

## Imagine the following simulated data

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.margin = TRUE)

```

```{r}

library(ggplot2) # beautiful graphs

library(ggmosaic) # mosaic plots

library(plotly) # animated graphs

library(tibble) # better data frames

library(pander) # nice tables

library(dplyr) # data wrangling

library(tidyr) # data wrangling

```

```{r}

hospital <- c(rep("better", 4),
              rep("normal", 4))

severity <- c(rep("less severe", 2),
              rep("more severe", 2),
              rep("less severe", 2),
              rep("more severe", 2))

outcome <- rep(c("success", "failure"), 4)

count <- c(18,2,
             32,48,
             64,16,
             4,16)

hospitaldata <- tibble::tibble(hospital, 
                               severity, 
                               outcome, 
                               count)

hospitaldatalong <- uncount(hospitaldata,
                                weights = count)

# hospitaldatalong # replay

```

```{r}

table(hospitaldatalong$outcome, 
      hospitaldatalong$hospital)

```

It appears that one is better off going to the *normal* hospital because there is a 68% success rate.

## Note that so far, we have a *simple* story.

We are only measuring the *outcome* of *failure* or *success* and the *type of hospital*.

```{r}

p1 <- ggplot(hospitaldatalong) + 
  # geom_mosaic(aes(x = product(outcome,
  #                             hospital), 
  #                 fill = outcome), na.rm=TRUE) +
  geom_bar(aes(x = hospital, 
               fill = outcome)) +
  labs(title = "Outcome by Type of Hospital",
       x = "type of hospital",
       y = "outcome")

ggplotly(p1)

```

## Let's add something to the story.

We now consider *severity of case*, in addition to *type of hospital* and *outcome*.

```{r}

table(hospitaldatalong$outcome, 
      hospitaldatalong$hospital, 
      hospitaldatalong$severity)

```

Considering severity of case, the rate of success is always better in the *better* hospital.

## Visualize it.

The disparity arises because more of the *more severe* cases go to the better hospital. <mark>Once we account for this disparity, we see the the *better* hospital performs better for both *levels of severity*.</mark>

```{r}

p2 <- ggplot(hospitaldatalong) + 
  # geom_mosaic(aes(x = product(outcome,
  #                             hospital), 
  #                 fill = outcome), na.rm=TRUE) +
  geom_bar(aes(x = hospital, 
               fill = outcome)) +
  facet_wrap(~severity, scales = "free") +
  labs(title = "Outcome by Type of Hospital AND Severity of Case",
       x = "type of hospital",
       y = "outcome")

ggplotly(p2)

```



