---
title: "Data Driven Document Demo"
author: "Andrew Grogan-Kaylor"
date: "today"
format:
  html:
    toc: true
    number-sections: true
    theme:
      light: yeti
      dark: vapor
  typst: 
    toc: true
    number-sections: true
    columns: 1
  docx: 
    toc: true
    number-sections: true
    reference-doc: markstat.docx
    fig-dpi: 600
    fig-height: 3
  epub: 
    toc: true
    number-sections: true
  revealjs: 
    output-file: data-driven-document-demo-revealjs.html
    transition: slide
    controls: true
    controls-tutorial: true
execute: 
  echo: false
  message: false
  warning: false
fig-height: 3
---


```{r}

library(gapminder) # gapminder data

library(ggplot2) # beautiful graphs

library(ggdist) # beautiful geoms

library(pander) # beautiful tables

```

# Background

Life expectancy differs in countries around the world. Life expectancy and *per capita* *Gross Domestic Product* (GDP) appear to be related.


```{r}

ggplot(gapminder,
       aes(x = lifeExp, 
           fill = continent)) +
  geom_histogram() +
  labs(title = "Distribution of Life Expectancy",
       x = "Life Expectancy") + 
  theme_minimal() +
  scale_fill_viridis_d() 

```

# Methods

Data were provided via the `gapminder` library which provides a sample of data available on [gapminder.org](http://www.gapminder.org)

# Results

Results indicate the per capita GDP and life expectancy are related in a non-linear way.

## Graph

```{r}

ggplot(gapminder,
       aes(x = gdpPercap,
           y = lifeExp,
           color = continent)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Life Expectancy by GDP",
       x = "GDP Per Capita",
       y = "Life Expectancy",
       caption = "Gapminder Data") +
  scale_color_viridis_d() +
  theme_minimal()

ggsave("scatterplot.png",
       last_plot() + theme(legend.position = "none"),
       width = 3, 
       height = 3)

```


## Table

```{r}

# pander(table(gapminder$continent),
#        caption = "Data Points by Continent")

pander(summary(gapminder$lifeExp),
       caption = "Life Expectancy")

pander(summary(gapminder$gdpPercap),
       caption = "GDP Per Capita")

```


# Discussion

*Per capita* GDP and life expectancy appear to be related. These findings may have implications for public health policy, for development policy, for intervention, and for advocacy.



