---
title: "How to Choose a Chart"
author: "Andy Grogan-Kaylor"
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    highlight: haddock
    toc: no
  pdf_document:
    highlight: haddock
    keep_tex: no
    latex_engine: xelatex
always_allow_html: yes
mainfont: Tw Cen MT
geometry: margin=0.75in
subtitle: A Statistically Motivated Guide
fontsize: 12pt
---

```{r global_options, echo=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.height = 2.0)

# NOTES: For HTML, echo=TRUE; for PDF, echo=FALSE

# Update:  As of 5-26-2017, the below is no longer necessary.

# Also, go through and change headers depending on format.
# \newpage just before "A Simulated Dataset"
#     Remove for HTML
#     Keep for PDF

```

```{r, call_libraries}

library(ggplot2) # beautiful graphs

library(ggthemes) # nice themes for ggplot2

library(ggbeeswarm) # helpful 'beeswarm' geometry

library(cowplot) # arrrange graphs

library(pander) # nice tables

```

```{r, eval=TRUE, echo=FALSE}

htmltools::tagList(rmarkdown::html_dependency_font_awesome())

```

```{r, eval = TRUE, echo = FALSE, fig.width = 7, fig.height = 2}

library(png)

library(grid)

banner <- readPNG("banner.png")

grid.raster(banner)

```

# How to Choose a Chart

Choosing the right chart to represent your data can be a daunting process. I believe that a *starting point* for this thinking is some basic statistical thinking about the *type* of variables that you have.  At the broadest level, variables may be conceptualized as *categorical* variables, or *continuous* variables.

* *categorical variables* represent unordered categories like *gender*, or *religious affiliation*.
* *continuous variables* represent a continuous scale like a *mental health scale*, or a *measure of neighborhood quality*.

Once we have discerned the type of variable that have, there are two followup questions we may ask before deciding upon a chart strategy:

* Is our graph about **one thing at a time**?
     + How much of *x* is there? 
     + What is the distribution of *x*?
* Is our graph about **two things at a time**?
     + What is the relationship of *x* and *y*?
     + How are *x* and *y* associated?

# A Few Notes

## A Note About Graph Labels

Graphs should have clear titles and labels.

```{r, echo=FALSE}

library(ggbeeswarm) # beeswarm plots

x <- rnorm(100, 100, 10)

ggplot(data = NULL,
       aes(x = x,
           y = 1,
           color = x)) +
  geom_beeswarm(groupOnX = FALSE) +
  geom_label(aes(x = 115, 
                 y = 1, 
                 label = "possibly \ninformative \nkey takeaway"),
             size = 2, 
             color = "red") +
  scale_colour_gradientn(colours=rainbow(7),
                         name = "some other \npiece of \ninformation") +
  ylab(" ") +
  xlab("informative label goes here") +
  labs(title = "Informative Title",
       subtitle = "Maybe The Title Is Even Your Main Takeaway",
       caption = "Words should be spelled out.  \nEvery extra 'dimension' (axes, color, shape) should convey information.") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 5))


```

<a id="software"></a>

## A Note About Software

The principles of graphing discussed in this document transcend any particular software package, and could be implemented in many different software packages, such as SPSS, SAS, Stata, or R.

The graphs in these particular examples use [ggplot2](http://ggplot2.org/), a graphing library in R.  ggplot2 graph syntax can be formidably complex, with a somewhat steep learning curve.  More information about ggplot can be found [here](http://www-personal.umich.edu/~agrogan/quick-intro-ggplot2/ggplot-in-3-easy-steps-maybe-2-easy-steps.html).

```{r, eval=FALSE}

# Note that ggplot2 can be MUCH simpler 
# than these examples make it look.
#
# For example,

  ggplot(mydata, aes(x = x)) + geom_histogram()

# will produce a perfectly serviceable histogram.
# 
# Much of the complication of the code in this document is simply
# the result of formatting tweaks to get the graphs EXACTLY
# the way I wanted them.

```

```{r, eval = FALSE}

# Note also, that for layout purposes, 
# I am reading each ggplot call 
# into an object, e.g.

p1 <- ggplot(mydata, aes(x = x)) + geom_histogram()

# so that I can later use plot_grid to lay out the graphs.

# In your own work, you do not need to do this, 
# and it may be simpler to simply say:

ggplot(...) + ...

```

<a id="colors"></a>

## A Note About Graph Colors

This document uses colors based upon official University of Michigan colors.  Using colors that match the design scheme of your organization may be helpful.

```{r michigan_colors}

# michigan colors

michigan_colors=c("#00274c", # blue
                  "#ffcb05", # maize
                  "#a4270b", # tappan red
                  "#e96300", # ross school orange
                  "#beb300", # wave field green
                  "#21c1bc", # taubman teal
                  "#2878ba", # arboretum blue
                  "#7207a5") # ann arbor amethyst

# name individual colors

michigan_blue <- "#00274c"

michigan_maize <- "#ffcb05"

tappan_red <- "#a4270b"

ross_school_orange <- "#e96300"

wave_field_green <- "#beb300"

taubman_teal <- "#21c1bc"

arboretum_blue <- "#2878ba"

ann_arbor_amethyst <- "#7207a5"

```

# A Simulated Data File of Continuous and Categorical Data

```{r simulate_data, echo = FALSE}

# simulated data

N <- 1000 # set sample size 

# categorical variables

# u <- rbinom(n = N, size = 1, prob = .75) 

u <- c(rep(0, N * .1),
       rep(1, N * .2),
       rep(0, N * .4),
       rep(1, N * .3)) # mixture of 0's and 1's

# v <- rbinom(n = N, size = 1, prob = .75)

v <- c(rep(0, N * .1),
       rep(1, N * .3),
       rep(0, N * .4),
       rep(1, N * .2)) # mixture of 0's and 1's

w <- rbinom(n = N, size = 1, prob = .25) 

# continuous variables

x <- rnorm(n = N, mean = 100, sd = 25) * (w + 1) # n, mean, sd

y <- rnorm(n = N, mean = 100, sd = 25) * (w + 1) # n, mean, sd

z <- rnorm(n = N, mean = 100, sd = 25) # n, mean, sd

# turn categorical data int factors

u <- factor(u, 
            levels = c(0, 1), 
            labels = c("Group A", "Group B"))

v <- factor(v, 
            levels = c(0, 1), 
            labels = c("Group A", "Group B"))

w <- factor(w, 
            levels = c(0, 1), 
            labels = c("Group A", "Group B"))

# new multinomial variable

s <- sample(c(1:4), 
            N,
            replace=TRUE, 
            prob=c(0.1, 0.5, 0.35, 0.05)) 

# new continuous variable

q <- x + 10 * s

# make s a factor

s <- factor(s,
            levels = c(1, 2, 3, 4),
            labels = c("Group 1", 
                       "Group 2", 
                       "Group 3", 
                       "Group 4"))

# read into data set

mydata <- data.frame(x, y, z, u, v, w, s, q)

# mydata

# save data to file

save(mydata, 
     file="how to choose a chart v3.RData")

```

A few randomly selected observations...

```{r, echo = FALSE}

mysample <- mydata[sample(1:nrow(mydata), 10, replace=FALSE),]

pander(mysample) # nice table

```

# One Thing At A Time \hspace{\stretch{1}} &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; Two Things At A Time

# Continuous \hspace{\stretch{1}} &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; Continuous By Categorical

```{r, eval=FALSE, echo=FALSE}

# Update 5-26-2017: Since HTML ignores Latex, and Latex ignores HTML
# the 2 syntaxes below can be combined, and this .Rmd file does not
# need to be modified based upon output type.

# HTML

# One Thing At A Time &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; Two Things At A Time

# Continuous &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; Continuous By Categorical

# PDF

# One Thing At A Time \hspace{\stretch{1}} Two Things At A Time

# Continuous \hspace{\stretch{1}} Continuous By Categorical

# 5-26-2017:  Another potential solution MIGHT be something like:

# | One Thing At A Time |Two Things At A Time |
# |:------|-----:|


```

<a id="histogram"></a>

```{r}

my_histogram <- ggplot(mydata, aes(x = x)) + 
  geom_histogram(fill = arboretum_blue) + 
  ggtitle("histogram") +
  xlab("continuous") + ylab("count") + 
  theme_minimal()

my_facet_histogram <- ggplot(mydata, aes(x = x)) + 
  geom_histogram(fill = arboretum_blue) + 
  facet_wrap(~w, nrow = 2) + 
  ggtitle("histogram by group") + 
  xlab("continuous") + ylab("count") + 
  theme_minimal() +
  theme(axis.text = element_text(size = 5)) # small font size for axis

plot_grid(my_histogram, my_facet_histogram, ncol=2)

```

<a id="density"></a>

```{r}

my_density <- ggplot(mydata, aes(x = y)) + 
  geom_density(fill = michigan_maize) + 
  ggtitle("density") +
  xlab("continuous") + ylab("density") + 
  theme_minimal()

my_facet_density <- ggplot(mydata, aes(x = y)) + 
  geom_density(fill = michigan_maize) + 
  facet_wrap(~w, nrow = 2) +
  ggtitle("density by group") +
  xlab("continuous") + ylab("density") + 
  theme_minimal() +
  theme(axis.text = element_text(size = 5)) # small font size for axis

plot_grid(my_density, my_facet_density, ncol = 2)

```

<a id="barchart_of_mean"></a>

```{r}

my_m_barchart <- ggplot(mydata, 
                        aes(x = 1, 
                            y = q, 
                            fill = factor(1))) + 
  stat_summary(fun.y = mean, geom = "bar") +
  scale_fill_manual(values = c(arboretum_blue)) + 
  ggtitle("barchart of mean") + 
  guides(fill=FALSE) +
  xlab(" ") + 
  ylab("mean of continuous") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x = element_blank())

my_facet_m_barchart <- ggplot(mydata, 
                              aes(x = factor(s), 
                                  y = q, 
                                  fill = s)) + 
  stat_summary(fun.y = mean, geom = "bar") + 
  scale_fill_manual(values = c(arboretum_blue, 
                               taubman_teal, 
                               michigan_blue,
                               michigan_maize)) + 
  ggtitle("barchart of mean \nby group") + 
  guides(fill=FALSE) +
  xlab("categorical") + ylab("mean of continuous") + 
  theme_minimal()

plot_grid(my_m_barchart, 
          my_facet_m_barchart, 
          ncol = 2)

```

```{r}

my_horiz_m_barchart <- ggplot(mydata, 
                              aes(x = 1, 
                                  y = q, 
                                  fill = factor(1))) + 
  stat_summary(fun.y = mean, 
               geom = "bar") +
  coord_flip() +
  scale_fill_manual(values = c(arboretum_blue)) + 
  ggtitle("horizontal barchart of mean") + 
  guides(fill=FALSE) +
  xlab(" ") + 
  ylab("mean of continuous") + 
  theme_minimal() +
  theme(axis.text.y = element_blank()) +
  theme(axis.ticks.y = element_blank())

my_facet_horiz_m_barchart <- ggplot(mydata, 
                                    aes(x = factor(s), 
                                        y = q, 
                                        fill = s)) + 
  stat_summary(fun.y = mean, 
               geom = "bar") +
  coord_flip() +
  scale_fill_manual(values = c(arboretum_blue, 
                               taubman_teal, 
                               michigan_blue,
                               michigan_maize)) +   
  ggtitle("horizontal barchart of mean \nby group") + 
  guides(fill=FALSE) +
  xlab(" ") + ylab("mean of continuous") + 
  theme_minimal() +
  theme(axis.text.y = element_blank()) +
  theme(axis.ticks.y = element_blank())

plot_grid(my_horiz_m_barchart,
          my_facet_horiz_m_barchart)

```

<a id="dotchart_of_mean"></a>

```{r}

my_horiz_m_dotchart <- ggplot(mydata, 
                              aes(x = 1, 
                                  y = q, 
                                  fill = factor(1))) + 
  stat_summary(fun.y = mean, 
               geom = "point", size = 5) +
  coord_flip() +
  scale_color_manual(values = c(arboretum_blue)) + 
  ggtitle("horizontal dotchart of mean") + 
  guides(fill = FALSE) +
  xlab(" ") + 
  ylab("mean of continuous") + 
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) 

my_facet_horiz_m_dotchart <- ggplot(mydata, 
                                    aes(x = factor(s), 
                                        y = q, 
                                        color = s)) + 
  stat_summary(fun.y = mean, 
               geom = "point", 
               size = 5) +
  coord_flip() +
  scale_color_manual(name = "group",
                     values = c(arboretum_blue, 
                                taubman_teal, 
                                michigan_blue,
                                michigan_maize)) +   
  ggtitle("horizontal dotchart of mean \nby group") + 
  guides(fill=FALSE) +
  xlab(" ") + 
  ylab("mean of continuous") + 
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.ticks = element_blank())

plot_grid(my_horiz_m_dotchart,
          my_facet_horiz_m_dotchart)

```

<a id="lollipop_chart_of_mean"></a>

```{r}

my_horiz_m_lollipop_chart <- ggplot(mydata, 
                                    aes(x = 1, 
                                        y = q, 
                                        fill = factor(1))) + 
  stat_summary(fun.y = mean, 
               geom = "point", 
               size = 5) +
  geom_segment(aes(x = 1,
                   xend = 1,
                   y = 0,
                   yend = mean(q))) +
  coord_flip() +
  scale_color_manual(values = c(arboretum_blue)) + 
  ggtitle("horizontal lollipop chart of mean") + 
  guides(fill = FALSE) +
  xlab(" ") + ylab("mean of continuous") + 
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) 

my_facet_horiz_m_lollipop_chart <- ggplot(mydata, 
                                    aes(x = factor(s), 
                                        y = q, 
                                        color = s)) + 
  stat_summary(fun.y = mean, 
               geom = "point", 
               size = 5) +
  geom_segment(aes(x = factor(s),
                   xend = factor(s),
                   y = 0,
                   yend = mean(q))) +
  coord_flip() +
  scale_color_manual(name = "group",
                     values = c(arboretum_blue, 
                                taubman_teal, 
                                michigan_blue,
                                michigan_maize)) +   
  ggtitle("horizontal lollipop chart of mean \nby group") + 
  guides(fill=FALSE) +
  xlab(" ") + 
  ylab("mean of continuous") + 
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.ticks = element_blank())

plot_grid(my_horiz_m_lollipop_chart,
          my_facet_horiz_m_lollipop_chart)

```


<a id="linechart_of_mean"></a>

```{r}

my_m_linechart <- ggplot(mydata, 
                         aes(x = factor(s), 
                             y = mean(q), 
                             group = 1)) + 
  stat_summary(fun.y = mean, 
               geom = "line", 
               size = 2, 
               color = arboretum_blue) +
  geom_blank() +
  ggtitle("linechart of mean") +
  xlab(" ") + 
  ylab("mean of continuous") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x = element_blank())

my_facet_m_linechart <- ggplot(mydata, 
                               aes(x = factor(s), 
                                   y = q, 
                                   group = 1)) + 
  stat_summary(fun.y = mean, 
               geom = "line", 
               size = 2, 
               color = arboretum_blue) +
  ggtitle("linechart of mean \nby group") + 
  xlab(" ") + 
  ylab("mean of continuous") + 
  theme_minimal() 

plot_grid(my_m_linechart, my_facet_m_linechart)
       
```

<a id="violin_plot"></a>

```{r}

my_violin <- ggplot(mydata, 
                    aes(x = 1, 
                        y = y)) + 
  geom_violin(fill = ross_school_orange) + 
  ggtitle("violin plot") + 
  xlab(" ") + 
  ylab("continuous") + 
  theme_minimal()  +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x = element_blank())

my_facet_violin <- ggplot(mydata, 
                          aes(x = 1, 
                              y = y)) + 
  geom_violin(fill = ross_school_orange) + 
  facet_wrap(~w, 
             ncol = 2) + 
  ggtitle("violin plot \nby group") + 
  xlab("categorical") + 
  ylab("continuous") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x = element_blank())

plot_grid(my_violin, my_facet_violin, ncol = 2)

```

<a id="boxplot"></a>

```{r}

my_boxplot <- ggplot(mydata, 
                     aes(x = 2, 
                         y = y)) + 
  geom_boxplot(colour=tappan_red) + 
  scale_x_discrete(limit = c(0,1,2)) +
  ggtitle("boxplot") + 
  xlab(" ") + 
  ylab("continuous") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x = element_blank())

my_conditional_boxplot <- ggplot(mydata, 
                                 aes(x = w, 
                                     y = y)) + 
  geom_boxplot(colour=tappan_red, width = .5) + 
  ggtitle("boxplot \nby group") + 
  xlab("categorical") + 
  ylab("continuous") + 
  theme_minimal()

plot_grid(my_boxplot, my_conditional_boxplot, ncol = 2)

```

<a id="beeswarm"></a>

```{r}

my_beeswarm <- ggplot(mydata, 
                      aes(x = y, 
                          y = 1)) + 
  geom_beeswarm(colour = ann_arbor_amethyst, 
                groupOnX = FALSE) + 
  ggtitle("beeswarm plot") + 
  xlab("continuous") + 
  ylab("") + 
  theme_minimal() +
  theme(axis.text=element_text(size = 5), 
        axis.text.y = element_blank()) 

my_facet_beeswarm <- ggplot(mydata, 
                            aes(x = y, 
                                y = 1)) + 
  geom_beeswarm(colour = ann_arbor_amethyst, 
                groupOnX = FALSE) + 
  facet_wrap(~w, nrow = 2) +
  ggtitle("beeswarm plot \nby group") + 
  xlab("continuous") + 
  ylab("") + 
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5),
        axis.text.y = element_blank()) 

plot_grid(my_beeswarm, my_facet_beeswarm)

```

```{r}

my_dotplot <- ggplot(mydata, 
                     aes(x = y)) + 
  geom_dotplot(colour=wave_field_green, 
               binwidth = 3.0) + 
  ggtitle("dotplot") + 
  xlab("continuous") + 
  ylab("density") + 
  theme_minimal()

my_facet_dotplot <- ggplot(mydata, 
                           aes(x = y)) + 
  geom_dotplot(colour=wave_field_green, 
               binwidth = 1.5) + 
  facet_wrap(~w, nrow = 2) + 
  ggtitle("dotplot \nby group") + 
  xlab("continuous") + 
  ylab("density") + 
  theme_minimal() +
  theme(axis.text = element_text(size = 5)) # small font size for axis

plot_grid(my_dotplot, 
          my_facet_dotplot, 
          ncol = 2)

```

# One Thing At A Time \hspace{\stretch{1}} &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; Two Things At A Time

# Categorical \hspace{\stretch{1}} &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; Categorical By Categorical

```{r, eval=FALSE, echo=FALSE}

#HTML

# One Thing At A Time &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; Two Things At A Time

# Categorical &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; Categorical By Categorical

# PDF

# One Thing At A Time \hspace{\stretch{1}} Two Things At A Time

# Categorical \hspace{\stretch{1}} Categorical By Categorical

```

```{r}

my_barchart <- ggplot(mydata, 
                      aes(s, 
                          fill = s)) + 
  geom_bar(width = 1.0) + 
  scale_fill_manual(values = c(arboretum_blue, 
                               taubman_teal,
                               michigan_blue,
                               michigan_maize)) + 
  ggtitle("bar chart") + 
  guides(fill=FALSE) +
  xlab("categorical") + 
  ylab("count") + 
  theme_minimal() 

my_facet_barchart <- ggplot(mydata, 
                            aes(s, 
                                fill = s)) + 
  geom_bar(width = 1.0) + 
  facet_wrap(~u, ncol = 2) + 
  scale_fill_manual(values = c(arboretum_blue, 
                               taubman_teal, 
                               michigan_blue,
                               michigan_maize)) + 
  ggtitle("bar chart \nby group") + 
  guides(fill=FALSE) +
  xlab("categorical") + ylab("count") + 
  theme_minimal() +
  theme(axis.text.x = element_text(size = rel(.6)))

plot_grid(my_barchart, my_facet_barchart, ncol = 2)

```


```{r}

my_horiz_barchat <- ggplot(mydata, aes(s, fill = s)) + 
  geom_bar(width = 1.0) + 
  coord_flip() +
  scale_fill_manual(values = c(arboretum_blue, 
                               taubman_teal, 
                               michigan_blue,
                               michigan_maize)) + 
  ggtitle("horizontal bar chart") + 
  guides(fill=FALSE) +
  xlab("categorical") + 
  ylab("count") + 
  theme_minimal() 

my_horiz_facet_barchart <- ggplot(mydata, aes(s, fill = s)) + 
  geom_bar(width = 1.0) + 
  facet_wrap(~u, ncol = 1) +
  coord_flip() +
  scale_fill_manual(values = c(arboretum_blue, 
                               taubman_teal, 
                               michigan_blue,
                               michigan_maize)) + 
  ggtitle("horizontal bar chart \nby group") + 
  guides(fill=FALSE) +
  xlab("categorical") + 
  ylab("count") + 
  theme_minimal() +
  theme(axis.text.y = element_text(size = rel(.5)))

plot_grid(my_horiz_barchat, my_horiz_facet_barchart, ncol = 2)

```


```{r}

my_pie <- ggplot(mydata, aes(x = factor(1), fill = v)) + 
  geom_bar(width = 1) + 
  coord_polar(theta="y") + 
  scale_fill_manual(values = michigan_colors) + 
  ggtitle("pie chart") + guides(fill=FALSE) +
  xlab(" ") + 
  ylab("categorical") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank()) + 
  theme(axis.ticks = element_blank())

my_facet_pie <- ggplot(mydata, aes(x = factor(1), fill = v)) + 
  geom_bar(width = 1, position = "fill") + 
  coord_polar(theta="y") + 
  facet_wrap(~u, ncol = 2) + 
  scale_fill_manual(values = michigan_colors) + 
  ggtitle("pie chart \nby group") + 
  guides(fill=FALSE) +
  xlab(" ") + 
  ylab("categorical") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank())  + 
  theme(axis.ticks = element_blank())

plot_grid(my_pie, my_facet_pie, ncol = 2)

```


```{r}

my_doughnut <- ggplot(mydata, aes(x = factor(1), fill = v)) + 
  geom_bar(width = .5) + 
  coord_polar(theta="y") + 
  scale_fill_manual(values = michigan_colors) + 
  ggtitle("doughnut chart") + 
  guides(fill=FALSE) +
  xlab(" ") + 
  ylab("categorical") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank()) + 
  theme(axis.ticks = element_blank())

my_facet_doughnut <- ggplot(mydata, aes(x = factor(1), fill = v)) + 
  geom_bar(width = .5, position = "fill") + 
  coord_polar(theta="y") + 
  facet_wrap(~u, ncol = 2) + 
  scale_fill_manual(values = michigan_colors) + 
  ggtitle("doughnut chart \nby group") + 
  guides(fill=FALSE) +
  xlab(" ") + 
  ylab("categorical") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank())  + 
  theme(axis.ticks = element_blank())

plot_grid(my_doughnut, my_facet_doughnut, ncol = 2)

```

# Continuous by Continuous

```{r}

my_scatterplot <- ggplot(mydata, aes(x=x, y=y)) + 
  geom_point(colour = ann_arbor_amethyst) + 
  ggtitle("scatterplot") + 
  xlab("continuous") + 
  ylab("continuous") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank())  + 
  theme(axis.ticks = element_blank())
  
my_scatterplot_smoother <- ggplot(mydata, aes(x=x, y=y)) + 
  geom_point(colour = ann_arbor_amethyst) + 
  geom_smooth(method = lm, color = michigan_maize, size = 2) + 
  ggtitle("scatterplot with fit line") + 
  xlab("continuous") + 
  ylab("continuous") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank())  + 
  theme(axis.ticks = element_blank())

plot_grid(my_scatterplot, my_scatterplot_smoother)

```

```{r}

my_hexagon <- ggplot(mydata, aes(x=x, y=y)) + 
  geom_hex() + 
  ggtitle("hexagon plot") + 
  xlab("continuous") + 
  ylab("continuous") + 
  theme_minimal() + 
  theme(legend.text = element_text(size=4), 
        legend.key.size = unit(.25, "cm")) + 
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank())  + 
  theme(axis.ticks = element_blank())  + 
  scale_fill_gradient(low = taubman_teal,
                      high = arboretum_blue) 

my_smoother <- ggplot(mydata, aes(x=x, y=y)) + 
  geom_point(colour = ann_arbor_amethyst) + 
  geom_smooth(se=TRUE, color=michigan_maize, size=2) + 
  ggtitle("scatterplot with smoother") + 
  xlab("continuous") + 
  ylab("continuous") + 
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank())  + 
  theme(axis.ticks = element_blank())

plot_grid(my_hexagon, my_smoother)

```


```{r}

my_area <- ggplot(mydata, aes(x=x, y=y)) + 
  geom_area(position = "stack", fill = ross_school_orange) +
  ggtitle("area plot") + 
  xlab("continuous") + 
  ylab("continuous") + 
  theme_minimal() + 
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank())  + 
  theme(axis.ticks = element_blank())

my_contour <- ggplot(mydata, aes(x=x, y=y)) + 
  stat_density_2d(aes(fill = ..level..), 
                  geom = "polygon") + 
  ggtitle("contour plot") + 
  xlab("continuous") + 
  ylab("continuous") + 
  theme_minimal() + 
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank())  + 
  theme(axis.ticks = element_blank()) + 
  scale_fill_gradient(low = taubman_teal,
                      high = arboretum_blue) 

plot_grid(my_area, my_contour)

```

```{r, eval=TRUE, echo=FALSE}

chartgrid <- plot_grid(my_histogram + 
                         theme(title=element_text(size = 20)), 
                       my_facet_histogram + 
                         theme(title=element_text(size = 20)), 
                       my_barchart + 
                         theme(title=element_text(size = 20)), 
                       my_facet_barchart + 
                         theme(title=element_text(size = 20)), 
                       my_scatterplot + 
                         theme(title=element_text(size = 20)),
                       my_density + 
                         theme(title=element_text(size = 20)),
                       my_facet_density + 
                         theme(title=element_text(size = 20)), 
                       my_horiz_barchat + 
                         theme(title=element_text(size = 20)),
                       my_horiz_facet_barchart + 
                         theme(title=element_text(size = 20)), 
                       my_scatterplot_smoother + 
                         theme(title=element_text(size = 20)), 
                       my_m_barchart + 
                         theme(title=element_text(size = 20)),
                       my_facet_m_barchart + 
                         theme(title=element_text(size = 20)), 
                       my_pie + 
                         theme(title=element_text(size = 20)), 
                       my_facet_pie + 
                         theme(title=element_text(size = 20)), 
                       my_contour + 
                         theme(title=element_text(size = 20)), 
                       my_beeswarm + 
                         theme(title=element_text(size = 20)), 
                       my_facet_beeswarm + 
                         theme(title=element_text(size = 20)), 
                       my_doughnut + 
                         theme(title=element_text(size = 20)),
                       my_facet_doughnut + 
                         theme(title=element_text(size = 20)), 
                       my_smoother + 
                         theme(title=element_text(size = 20)), 
                       ncol = 5) 

save_plot("chartgrid.png", 
          chartgrid,
          ncol = 5, 
          nrow = 4, 
          base_aspect_ratio = 1.25)

# Banner

banner <- plot_grid(my_histogram + 
                      theme(title=element_text(size = 20)), 
                    my_density + 
                      theme(title=element_text(size = 20)), 
                    my_pie + 
                      theme(title=element_text(size = 20)), 
                    my_scatterplot_smoother + 
                      theme(title=element_text(size = 20)), 
                    my_beeswarm + 
                      theme(title=element_text(size = 20)), 
                    nrow = 1)

save_plot("banner.png", banner,
          ncol = 5,
          nrow = 1)

save_plot("h.png",
          my_histogram + 
            theme(title = element_text(size = 20)),
          base_height = 4)

save_plot("b.png",
          my_barchart + 
            theme(title = element_text(size = 20)),
          base_height = 4)

save_plot("s.png",
          my_scatterplot_smoother + 
            theme(title = element_text(size = 20)),
          base_height = 4)



```


Graphics made with the [ggplot2](http://ggplot2.org/) graphing library created by Hadley Wickham.

Available online at [https://agrogan1.github.io/](https://agrogan1.github.io/)

*How to Choose a Chart* by Andrew Grogan-Kaylor is licensed under a [Creative Commons Attribution-ShareAlike 4.0 International License](http://creativecommons.org/licenses/by-sa/4.0/).  You are welcome to download and use this handout in your own classes, or work, as long as the handout remains properly attributed.

Last updated: ```r format(Sys.Date(),"%B %d %Y")```  at ```r format(Sys.time(),"%H:%M")```



