---
title: "Variables & Visualization"
author: "Andy Grogan-Kaylor"
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
output:
  revealjs::revealjs_presentation:
    fig_height: 4
    highlight: haddock
    reveal_options:
      chalkboard:
        theme: chalkboard
        toggleChalkboardButton: yes
        toggleNotesButton: yes
      previewLinks: yes
      progress: yes
      slideNumber: yes
    reveal_plugins:
    - chalkboard
    - zoom
    self_contained: no
    theme: night
    transition: convex
  tufte::tufte_handout:
    fig_height: 2.5
    toc: yes
    number_sections: yes
    latex_engine: xelatex
subtitle: What Is The Story You Are Trying To Tell?
---

```{r setup, include=FALSE}

# invalidate cache when the tufte version changes

knitr::opts_chunk$set(tidy = FALSE, 
                      echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.margin = TRUE,
                      cache.extra = packageVersion('tufte'))

options(htmltools.dir.version = FALSE)

options(scipen = 999)

library(tufte)

library(png) # png

library(grid)

# library(WDI) # for accessing World Bank data

library(ggplot2)

library(ggbeeswarm)

library(ggthemes)

library(viridis)

library(ggpomological)

library(ggmap) # maps

library(cowplot)

library(psych)

library(pander) # nice tables

library(maps)

library(choroplethr) # library for making maps

library(choroplethrMaps) # map data

```

```{r}

N <- 100

age <- rnorm(N, 40, 10)

mental_health <- rnorm(N, 100, 15)

income <- rlnorm(N, 0, 1) * 10000 + 10000

group <- rbinom(N, 1, .2)

group <- factor(group,
                levels = c(0,1),
                labels = c("Group A", "Group B"))

neighborhood <- cut(runif(N),
              c(0, 0.45, 0.90, 1), labels=FALSE)

neighborhood <- factor(neighborhood,
                 levels = c(1,2,3),
                 labels = c("Neighborhood A", 
                            "Neighborhood B", 
                            "Neighborhood C"))

mydata <- data.frame(age, mental_health, group, neighborhood)

```

# How to Navigate This Presentation (scroll down &#8681;) 

## Navigation

* **o** for outline
* **f** for full screen
* **alt-click** for zoom
* **pen** icons in lower left hand corner can be used for annotations.

# Introduction (scroll down &#8681;) 

## Possibilities

```{r, eval = TRUE, fig.margin=FALSE, fig.cap="possible visualizations"}

img <- readPNG("banner.png")

grid.raster(img)

```

## Background

* Deciding upon the right data visualization to represent your data can be a daunting process.
* I believe that a *starting point* for this thinking is some basic statistical thinking about the *type* of variables that you have. 
* At the broadest level, variables may be conceptualized as *categorical* variables, or *continuous* variables.

## Data Often Come From A Survey Questionnaire.

```{r, eval = TRUE, fig.margin=TRUE, fig.height=3, fig.cap="hypothetical questionnaire"}

img <- readPNG("questionnaire.png")

grid.raster(img)

```

## What is Data? 

A data set is nothing more than a series of rows and columns that contain
answers to responses to a survey.

* Rows are usually used for individuals, while columns indicate the
questionnaire answers, or measures, from those people.
* Answers to questions are often given numerical responses (e.g. "no"
is frequently coded as "0" and "yes" is frequently coded as "1")

```{r}

person <- c(1,2,3)

Q1 <- c(1,2,1)

Q2 <- c(0,0,1)

Q3 <- c(100, 200, -9)

sample_dataset <- data.frame(person, Q1, Q2, Q3)

pander(sample_dataset, 
       caption = "Hypothetical Data")

# sample_dataset

```


## Some Notes on Data

* In working through our research questions, we'll constantly be going back and forth between the actual data (to see the pattern of responses) and the documentation, to figure out the actual question asked as well as how the different responses are coded.
* Often in a spreadsheet, you'll see the full text of a question written out (e.g. "What is your gender identity"?)
* Most programs that work with data are going to want abbreviations (e.g. "Q1" or "gender"") for the questions. These abbreviations should usually have no spaces and be 8 characters or less.

## Missing Data

* One cell of the sample data set has a negative number.
* Frequently negative numbers are used to indicate what are called "missing values". A missing value is a response like "don't know"" or "refused to answer" or "did not answer". 
* Before we start doing calculations with our data, we'll want to change negative numbers to true missing values (usually symbolized by a "." or "NA", so that they don't goof up our calculations.)

## What are Variables?

* By variables, I simply mean the columns of data that you have.
* For our purposes, you may think of variables as synonymous with questionnaire items, or columns of data.

## Variable Types

* *categorical variables* represent unordered categories like *neighborhood*, or *religious affiliation*, or *place of residence*.
* *continuous variables* represent a continuous scale like a *mental health scale*, or a *measure of life expectancy*.

## A Data Visualization Strategy

Once we have discerned the type of variable that have, there are two followup questions we may ask before deciding upon a chart strategy:

* Is our graph about **one thing at a time**?
     + How much of *x* is there? 
     + What is the distribution of *x*?
* Is our graph about **two things at a time**?
     + What is the relationship of *x* and *y*?
     + How are *x* and *y* associated?

## More On Strategy

```{r}

x <- rnorm(1000, 100, 10)

```

```{r, eval=TRUE, fig.margin = TRUE, fig.fullwidth = FALSE}

ggplot(data = NULL,
       aes(x = x,
           y = 1,
           color = x)) +
  geom_beeswarm(groupOnX = FALSE) +
  geom_label(aes(x = 120, 
                 y = 1, 
                 label = "possibly \ninformative \nkey takeaway"),
             size = 2, 
             color = "red") +
  scale_color_gradient(name = "some other \npiece of \ninformation",
                       low = "#375E97",
                       high = "#FFBB00") +
  ylab("possibly another dimension of the data") +
  xlab("one dimension of the data") +
  labs(title = "Informative Title",
       subtitle = "Maybe The Title Is Even Your Main Takeaway",
       caption = "Every extra 'dimension' (x axis, y axis, color, shape) 
       should convey information.
       \n Words should be spelled out.") + 
  theme_minimal()

```

# Simulated Data (scroll down &#8681;)

## Background

This example uses simulated data on social work clients, of the kind that a social service agency might collect.

```{r, fig.margin=TRUE, eval=FALSE, fig.cap="Descriptive Statistics"}

pander(describe(mydata, 
                skew = FALSE,
                ranges = FALSE),
       caption = "Descriptive Statistics")

```

```{r, eval=TRUE, fig.cap="First Few Lines of Data File"}

# head(mydata)

pander(head(mydata, 5),
       caption = "Simulated Data")

```

# Show One Thing At A Time. (scroll down &#8681;)

## Rationale

We start by visualizing one indicator at a time.

## Continuous Variable

Sometimes the most interesting visualizations, are visualizations that give us a sense of the maximum, minimum, and average values.  For example, the *histogram* and *beeswarm plot* display information on *age*.

```{r, fig.cap="Histogram and Beeswarm Plot"}

p1 <- ggplot(mydata, 
             aes(x = age)) +
  labs(title = "Distribution of Age") +
  theme_minimal()
  
p1a <- p1 + 
  geom_histogram(fill = "#375E97") 

# p1a

p1b <- p1 + 
  geom_beeswarm(aes(y = 1), 
                groupOnX = FALSE,
                cex = 2.0,
                color = "#375E97") +
  ylab("") +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

# p1b 

plot_grid(p1a, p1b)

# ggsave("p1.png", p1)

```


## Categorical Variable

We would use a slightly different visualization, for example, a barchart, when our data are grouped into categories.

```{r, fig.cap="Barchart"}

p2 <- ggplot(mydata,
             aes(x = neighborhood,
                 fill = neighborhood)) +
  geom_bar() + 
  labs(title = "Neighborhood") +
  # scale_fill_viridis(discrete = TRUE,
  #                    option = "plasma") +
  scale_fill_canva(palette = "Primary colors with a vibrant twist") +
  theme_minimal()

p2

```

# Show The Relationship Of Two Things. (scroll down &#8681;)

## Rationale

Our task becomes somewhat more complicated when we want to understand the relationship of *one thing* to *another thing.*

## Categorical by Categorical

Here, for example, we visualize two *categorical* variables, *neighborhood*, by *group*. In this graph, the height of the bars represents the *count* of observations.

```{r, fig.cap="Barchart"}

p3 <- ggplot(mydata,
             aes(x = neighborhood, 
                 fill = neighborhood)) +
  geom_bar() +
  facet_wrap(~group) +
  coord_flip() + 
  labs(title = "Neighborhood",
       subtitle = "by Group") +
  # scale_fill_viridis(discrete = TRUE,
  #                    option = "plasma") +
  scale_fill_canva(palette = "Primary colors with a vibrant twist") +
  theme_minimal() +
  theme(strip.text = element_text(size = 8))

p3

```

## Continuous by Continuous

Here, we visualize two *continuous* variables, *mental health*, by *age*. 

```{r, fig.cap="Scatterplot"}

p4 <- ggplot(mydata,
             aes(x = age, y = mental_health)) + 
  geom_point(color = "#375E97") +
  geom_smooth(color = "#FB6542") +
  labs(title = "Mental Health by Age") +
  ylab("mental health")  +
  theme_minimal()

p4

```

## Continuous by Categorical

Last, we visualize a *continuous* variable by a categorical variable, *mental health*, by *group*. In this graph, the height of the bars represents the *mean score*.

```{r, fig.cap="BarChart"}

p5 <- ggplot(mydata,
             aes(x = group,
                 y = mental_health,
                 fill = group)) +
    stat_summary(fun.y = mean, 
                 geom = "bar") + 
  coord_flip() +
  labs(title = "Mental Health by Group") + 
  ylab("mental health") +
  # scale_fill_viridis(discrete = TRUE,
  #                    option = "plasma") +
  scale_fill_canva(palette = "Primary colors with a vibrant twist") +
  theme_minimal()

p5

```

```{r, eval = FALSE}

# Show a process

```

# Show Where Something Is. (scroll down &#8681;)

## Rationale

Sometimes our task is different. We want to visualize information, but add information on spatial location, using a map.

## Map

```{r, fig.cap="Map", fig.height=4}

world <- map_data("world")

world$myfill <- "no clients"

world$myfill[world$region == "Canada"] <- "clients"

world$myfill[world$region == "USA"] <- "clients"

world$myfill[world$region == "Mexico"] <- "clients"

mymap <- ggplot() + 
  geom_polygon(data = world, aes(x=long, 
                                 y = lat, 
                                 group = group,
                                 fill = myfill),
               color = "grey") + 
  labs(title = "Location of Agency Clients") +
  # scale_fill_viridis(discrete = TRUE,
  #                    option = "plasma") +
  scale_fill_canva(palette = "Primary colors with a vibrant twist") +
  theme_minimal()  + 
  coord_fixed(1.3) 

mymap

```

# Credits (scroll down &#8681;)

## ggplot2

Graphics made with the [ggplot2](http://ggplot2.org/) graphing library created by Hadley Wickham.

## More Information

* A more exhaustive set of graphical possibilities, along with the code to generate them in [R](https://www.r-project.org/) and [ggplot2](http://ggplot2.org/), can be found at my tutorial on [How To Choose A Chart](www.umich.edu/~agrogan/how_to_choose_a_chart).
* I have also written an [Introduction to R](http://www-personal.umich.edu/~agrogan/R/introduction-to-R.html) 

## Author

*Variables & Visualization* by [Andrew Grogan-Kaylor](https://agroganweb.wordpress.com/) is licensed under a [Creative Commons Attribution-ShareAlike 4.0 International License](http://creativecommons.org/licenses/by-sa/4.0/).  You are welcome to download and use this handout in your own classes, or work, as long as the handout remains properly attributed.

Last updated: ```r format(Sys.Date(),"%B %d %Y")```  at ```r format(Sys.time(),"%H:%M")```

```{r make_banner}

banner <- plot_grid(p1b + 
                      theme(plot.title=element_text(size = 20)), 
                    p2 + 
                      theme(plot.title=element_text(size = 20),
                            axis.text.x = element_blank()), 
                    p3 + 
                      theme(plot.title=element_text(size = 20)), 
                    p4 + 
                      theme(plot.title=element_text(size = 20)), 
                    p5 + 
                      theme(plot.title=element_text(size = 20)), 
                    nrow = 2,
                    mymap)

save_plot("banner.png", 
          banner,
          ncol = 3,
          nrow = 2)

```









