---
title: "Advanced `ggplot`"
author: "Andy Grogan-Kaylor"
date: "`r Sys.Date()`"
output: 
  slidy_presentation: 
    fig_caption: yes
    incremental: yes
    highlight: haddock
    css: ../UMslidy.css
    fig_height: 4
    fig_retina: 3
    slide_level: 2
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(emo)

```

# Navigation

* This is a slide deck. You may need to **scroll down** to see everything on longer slides.
* Use the `r emo::ji("left_arrow")` `r emo::ji("right_arrow")` keys to move through the presentation.
* Press `b` to make text *bigger*. Press `s` to make text *smaller*.

# Simulated Data From Social Service Agency

These are *simulated* data designed to be similar to the data that might come from a social service agency.

The data contain the following (*simulated*) variables:  `ID`, `age`, `gender`, `race_ethnicity`, `family_income`, `program`, `mental_health_T1`, `mental_health_T2`, `latitude`, `longitude`.

The *mental health* variables are scaled to have an average of 100. Lower numbers indicate lower mental health, while higher numbers indicate higher mental health.

There are some differences in *mental health* status in the data and an interesting exercise could be to use software like Excel, Google Sheets, Tableau or R to try to see which factors predict these differences.

```{r, echo=FALSE, eval=FALSE}

load("social-service-agency.RData")

clients$neighborhood <- sample(c("Neighborhood A",
                                 "Neighborhood B",
                                 "Neighborhood C"), 
                               453, 
                               replace = TRUE)

save(clients, file = "clients.RData")

```

# The Data

```{r}

load("clients.RData")

```

```{r, echo=FALSE}

DT::datatable(clients,
              extensions = 'Buttons', 
              options = list(dom = 'Bfrtip', 
                             buttons = c('copy', 
                                         'csv', 
                                         'excel')))

```


# Review of Basic `ggplot`

1. data
2. `aes`thetic
3. `geom`etry

```{r}

library(ggplot2) # call the library

```

```{r}

ggplot(clients,
       aes(x = program,
           y = mental_health_T2)) +
  stat_summary(fun = mean, # summarizing y 
               geom = "bar") # with bars

```

# Add A Theme

```{r}

ggplot(clients,
       aes(x = program,
           y = mental_health_T2)) +
  stat_summary(fun = mean, # summarizing y 
               geom = "bar") + # with bars
  theme_minimal()

```

# Alternate Geometries {#majorsection}

## Points

```{r}

ggplot(clients,
       aes(x = program,
           y = mental_health_T2)) +
  geom_point() +
  theme_minimal()

```

## Jittered Points

```{r}

ggplot(clients,
       aes(x = program,
           y = mental_health_T2)) +
  geom_jitter() +
  theme_minimal()

```

## Dotplots

```{r}

# install.packages("ggdist")

library(ggdist) # distribution plots

ggplot(clients,
       aes(x = program,
           y = mental_health_T2)) +
  stat_dots() + # dotplot geometry
  theme_minimal()

```

## Boxplots

```{r}

ggplot(clients,
       aes(x = program,
           y = mental_health_T2)) +
  geom_boxplot() + # boxplot geometry
  theme_minimal()

```


# Use of Color {#majorsection}

## Color As An *Aesthetic Element*

> e.g. *U.N. Blue*. 

Here the use of color adds *aesthetic appeal*. We do this by placing `fill` and `color` in the `geom`etry.

```{r}

ggplot(clients,
       aes(x = program,
           y = mental_health_T2)) +
  stat_dots(fill = "#009edb", # fill color
            color = "black") + # outline color
  theme_minimal()

```

## Color As Information

> Compare the minimal and maximal philosophies.

Here, we place `fill` color in the `aes`thetic so that color adds additional information, i.e. the gender of respondents.

(We still retain `color` in the `geom` to govern *outline* color)

```{r}

ggplot(clients,
       aes(x = program,
           fill = gender, # color as `information`
           y = mental_health_T2)) +
  stat_dots(color = "black") + # dotplot geometry
  theme_minimal()

```

## Color Palettes

> `viridis` color palette

```{r}

ggplot(clients,
       aes(x = program,
           fill = gender, # color as `information`
           y = mental_health_T2)) +
  stat_dots(color = "black") + # dotplot geometry
  scale_fill_viridis_d() + # beautiful colors
  theme_minimal()

```

# Coordinate Systems {#majorsection}

## Flipping The Coordinates

```{r}

ggplot(clients,
       aes(x = program,
           fill = gender, # color as `information`
           y = mental_health_T2)) +
  stat_dots(color = "black") + # dotplot geometry
  scale_fill_viridis_d() + # beautiful colors
  coord_flip() + # flip the coordinates
  theme_minimal()

```

# Small Multiples {#majorsection}

# Using `facet_wrap()`

```{r}

ggplot(clients,
       aes(x = program,
           fill = gender, # color as `information`
           y = mental_health_T2)) +
  stat_dots(color = "black") + # dotplot geometry
  scale_fill_viridis_d() + # beautiful colors
  coord_flip() + # flip the coordinates
  facet_wrap(~neighborhood) + # facet on neighborhood
  theme_minimal()

```

# Titles & Labels {#majorsection}

## Adding Titles and Labels

```{r, fig.height=6}

ggplot(clients,
       aes(x = program,
           fill = gender, # color as `information`
           y = mental_health_T2)) +
  stat_dots(color = "black") + # dotplot geometry
  scale_fill_viridis_d() +
  coord_flip() +
  labs(title = "Program Enrollment",
       subtitle = "By Gender Identity",
       caption = "Higher Mental Health Scores Are Better",
       y = "Mental Health At Time 2",
       x = "Program") +
  theme_minimal() +
  theme(plot.title = element_text(size = rel(1.5),
                             color = "darkblue"))

```

# Questions? {#majorsection}